EBASE manuscript

### 2023-01-05 manu-draft.qmd
more on methods

where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respectively, modeled as a function of the change in dissolved oxygen ($\delta{C_d}$) per unit time of the input data ($\delta{t}$).  Five unknown parameters are estimated by the model: $a$, $p$, $r$, $\Theta$, and $K$.  The required inputs are the time series of surface irradiance ($PAR$, $\mu$mol/m$^2$/s), water temperature ($T$, C), diel oxygen ($C_d$, mmol/m$^2$), and oxygen saturation ($C_{sat}$, mmol/m$^3$). Dissolved oxygen at saturation $C_{sat}$ is calculated from the input data as a function of water temperature, salinity, and barometric pressure.

@@

Instead of one of the traditional wind-based parameterization used in oceanography [e.g., @Wanninkhof14], @Grace15 uses the following for the gas transfer:

@@

where $C_O$ is solubility of dissolved oxygen per volume of seawater at the given temperature and $\rho$ (kg/m$^3$) is seawater density at zero elevation given temperature and salinity [@Millero81]. $C_O$ is calculated using a polynomial equation:

@@

The second modification to @eq-base for EBASE is a simplification to remove the $p$ and $\Theta$ parameters from the production (P) and respiration (R) components, respectively, of the mass balance equation.  The BASEmetab method provides options to estimate both $\Theta$ and $p$ given the data or to set as fixed values defined in the model file ($\Theta = 1.07177$, $p=1$). For EBASE, the temperature dependency of respiration is removed by setting $\Theta^{(T-\bar{T})}$ = 1.  The $p$ parameter, which is dimensionless, is also set to 1 as in @Holtgrieve10. Removing these components from the model improved model stability by creating more reasonable estimates for the other parameters.  This further simplifies @eq-basewan to:

@@

As for BASEmetab, the metabolic estimates in EBASE are defined by the change in dissolved oxygen over the time step $\delta{C_d}/\delta{t}$, where gross production is provided by $aPAR$, respiration is provided by $r$, and gas exchange is provided by the remainder [as in @Wanninkhof14].  Units for each of the metabolic estimates are $\runi$.

The remaining three parameters $a$ (light efficiency, $\auni$), $r$ (respiration, $\runi$), and $b$ ($\buni$) are estimated by likelihood given the observed data with the JAGS model using prior distributions shown in the model file. At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses, and then a finite difference approximation is used to estimate modeled oxygen concentration.  The estimated dissolved oxygen concentration is also returned, which can be compared to observed as one measure of model performance.

@@

The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each.  As for the BASEmetab model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can of course be changed based on previous knowledge of parameter specific to an ecosystem.  The default priors were based on those that returned reasonable parameter estimates using a forward-feeding oxygen mass balance equation call FWOXY (ForWard OXYgen).  For FWOXY, a dissolved oxygen time series is simulated by providing known parameter values in an equation similar to @eq-ebase. All of the physical parameters are held constant, e.g., wind speed, temperature, and salinity.  The other required inputs are the $a$, $r$, and $b$ parameters, which are set by the user.  As such, the dissolved oxygen output simulated by FWOXY as change per unit time is modelled by EBASE using the constant physical parameters as input. Reasonable prior distributions by EBASE were identified (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]) based on the ability to return the known parameters from FWOXY:

@@

The $T$ notation is JAGS model syntax indicating the distributions are truncated to a minimum (left) and maximum (right) value defined in the parentheses.  All prior distributions were constrained to positive values based on requirements of the model formula in @eq-ebase.  Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.504 (~twice the default mean).  @Wanninkhof14 states that the $b$ parameter typically does not vary by more than 20% and the standard deviation of the default prior is sufficiently high to expand beyond this range. Initial development of EBASE also showed that using an undefined upper limit of $b$ led to unstable and unreasonable parameter estimates. The density distributions for the default priors in EBASE are shown in @fig-priorplot.

@@

For ease of application, an R package was developed to implement the EBASE methods described above. A detailed web page (<https://fawda123.github.io/EBASE/>) describes the required data inputs for estimating the metabolic parameters using the core R functions with the JAGS software.  A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed.  Users can also specify alternative prior distributions from those in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] for each parameter. Additional functions are available for viewing model results, including a plot of the modelled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters. All results provided herein were generated using the EBASE R package.

@@

* Comparison with Odum and BASEmetab: Fwoxy Appalachicola

### Sensitivity analyses

Several configurations of the EBASE model can influence how the metabolic rates are estimated.  A simulated time series with known values for the $a$, $r$, and $b$ parameters that are returned by EBASE was used to determine if 1) EBASE estimates similar values, and 2) how the estimated values change with different model configurations.  Two types of sensitivity analyses were conducted to evaluate changes in the parameters returned by EBASE, where in all cases, the parameters were compared to the known values in the simulated time series. First, parameter estimates were compared to the known values after changing characteristics of their prior distributions.  Second, parameter estimates were compared using a model period of optimization for one day and one week of observations.  Details of these analyses are as follows.

@@

The metabolic estimates returned from each analysis were compared to the known values to identify priors the model was most sensitive to and if the conclusions varied by optimization period.  Estimates for the modeled dissolved oxygen, gross production, respiration ($r$), gas exchange ($D$), and $a$ were compared to those from the simulated time series using coefficient of variation ($R^2$) and root mean square error ($RMSE$) summaries of the 1:1 comparison of the estimated vs simulated values.  For the one day optimization period, all parameter estimates returned at the time step of the input data (i.e., modeled dissoled oxygen, $D$, and $a$) were daily-averaged for comparison.  For the seven day optimization period, all parameter estimates and those from the known time series were weekly-averaged.  Estimates for the $b$ parameter were evaluated only by visual comparison because a single value was used in the simulated time series, preventing the estimate of quantitative comparisons.

@@

Metabolic estimates from EBASE were compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach. An approximate one year continuous time series from Apalachicola Bay, Florida was provided as input to each method.  The input data were 2012 observations from February 22nd to December 19th at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, -84.8802$^{\circ}$ W) and included 15 minute observations of dissolved oxygen (mg/L), water temperature (C), salinity (psu), and sonde depth (m).  Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, ~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), PAR (mmol/m$^2$ as total per 15 minute observation), and barometric pressure (mb).  The combined input data were provided to each method after converting the units as required (e.g., BASEmetab requires PAR as $\mu$mol m$^{-2}$ s$^{-1}$).  Water column depth at Cat Point is approximately 1.8 m and all daily metabolic estimates returned by each method were converted to volumetric rates by dividing by the depth ($\mu$mol O$_2$ m$^{-3}$ d$^{-1}$).  Each comparison also evaluated metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15].  The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection.  All metabolic estimates were obtained using the default settings for each of the three methods, with an exception that the prior distribution for the reaeration coefficient ($K$) in BASEmetab was informed by results from the Odum output.  Initial assessments indicated that unstable and inaccurate results are obtained from BASEmetab using an uninformed prior distribution for $K$.  All other prior distributions for BASEmetab and EBASE were set as the software default.  A one-day model optimization period was used for EBASE.

@@

### Application to represenatiave NERR sites

>>>>

where the terms in the square brackets are production (P), respiration (R), and gas exchange (D), respectively, modeled as a function of the change in dissolved oxygen ($\delta{C_d}$) per unit time of the input data ($\delta{t}$).  Five unknown parameters are estimated by the model: $a$, $p$, $r$, $\Theta$, and $K$.  The required inputs are the time series of surface irradiance ($PAR$, $\mu$mol/m$^2$/s), water temperature ($T$, C), atmospheric pressure (atm), diel oxygen ($C_d$, mmol/m$^2$), and oxygen saturation ($C_{sat}$, mmol/m$^3$). Dissolved oxygen at saturation $C_{sat}$ is calculated from the input data as a function of water temperature, salinity, and barometric pressure.  Several changes, outlined and justified below, were made to @eq-base for EBASE.

@@

Instead of one of the traditional wind-based parameterization used in oceanography [e.g., @Wanninkhof14], @Grace15 uses the following for the gas transfer in BASEmetab:

@@

where $C_O$ is thr solubility of dissolved oxygen per volume of seawater at the given temperature and $\rho$ (kg/m$^3$) is seawater density at zero elevation given temperature and salinity [@Millero81]. $C_O$ is calculated using a polynomial equation:

@@

The second modification to @eq-base for EBASE is a simplification to remove the $p$ and $\Theta$ parameters from the production (P) and respiration (R) components, respectively, of the mass balance equation.  The BASEmetab method provides options to estimate both $\Theta$ and $p$ from the data or to set as fixed values defined in the model file ($\Theta = 1.07177$, $p=1$). For EBASE, the temperature dependency of respiration is removed by setting $\Theta^{(T-\bar{T})}$ = 1.  The $p$ parameter, which is dimensionless, is also set to 1 as in @Holtgrieve10. Removing these components mproved model stability by creating more reasonable estimates for the other parameters.  This further simplifies @eq-basewan to the final form for EBASE:

@@

As for BASEmetab, the metabolic estimates in EBASE are defined by the change in dissolved oxygen over the time step, $\delta{C_d}/\delta{t}$, where gross production is provided by $aPAR$, respiration is provided by $r$, and gas exchange is provided by the remainder [as in @Wanninkhof14].  Units for each of the metabolic estimates are $\runi$.

The remaining three parameters $a$ (light efficiency, $\auni$), $r$ (respiration, $\runi$), and $b$ ($\buni$) are estimated by likelihood given the observed data and prior distributions for the parameters.  The JAGS software [@Plummer03] is used with the EBASE R package [described below, @Beck22;@RCT22] to estimate the unknown parameters using Markov Chain Monte Carlo simulations in a Bayesian framework. At each time step, the change in oxygen concentration between time steps is calculated from the equation using model inputs and parameter guesses, and then a finite difference approximation is used to estimate modeled oxygen concentration. Estimations proceed using a forward in time integration where the dissolved oxygen at each time step is informed by parameter estimates from the preceding time steps.  The estimated dissolved oxygen concentration is also returned, which can be compared to observed as one measure of model performance.

@@

The metabolism estimates and their parameters returned by the Bayesian routine implemented in JAGS are affected by the prior distributions assigned to each.  As for the BASEmetab model, relatively uninformed prior distributions following a normal Gaussian distribution are used by default, although the priors can of course be changed based on previous knowledge of parameters specific to an ecosystem or as informed by other metabolic modeling approaches.  The default priors were based on those that returned reasonable parameter estimates using a forward-feeding oxygen mass balance equation call FWOXY [ForWard OXYgen, @Arriola22].  For FWOXY, a dissolved oxygen time series is simulated by providing known parameter values in an equation similar to @eq-ebase. All of the physical parameters are held constant, e.g., wind speed, temperature, and salinity.  The other required inputs are the $a$, $r$, and $b$ parameters, which are set by the user.  An initial oxygen concentration is provided and the simulated time series is created by forward addition of the inputs. As such, the dissolved oxygen output simulated by FWOXY as change per unit time were modelled by EBASE using the constant physical parameters as input. Reasonable prior distributions by EBASE were identified (Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior]) based on the ability to return the known parameters from FWOXY:

@@

The $T$ notation is JAGS model syntax indicating the distributions are truncated to a minimum (left) and maximum (right) value defined in the parentheses.  All prior distributions were constrained to positive values based on known physical constraints and requirements of the model formula in @eq-ebase.  Additionally, the prior distribution for $b$ was constrained to an upper limit of 0.504 (~twice the default mean).  @Wanninkhof14 states that the $b$ parameter typically does not vary by more than 20% and the standard deviation of the default prior is sufficiently high to expand beyond this range. Initial development of EBASE also showed that using an undefined upper limit of $b$ led to unstable and unreasonable parameter estimates. The density distributions for the default priors in EBASE are shown in @fig-priorplot.

@@

For ease of application, an R package was developed to implement the EBASE methods described above [@Beck22]. A detailed web page (<https://fawda123.github.io/EBASE/>) describes the required data inputs for estimating the metabolic parameters using the core R functions that pass inputs to the JAGS software.  A JAGS model file is included with the package that implements @eq-ebase, with options to supply a custom model file as needed.  Users can also specify alternative prior distributions from those in Equations [-@eq-aprior], [-@eq-rprior], and [-@eq-bprior] for each parameter. Additional functions are available for viewing model results, including a plot of the modelled dissolved oxygen with the observed, a time series plot of the primary metabolic estimates (P, R, and D), and a time series plot of the credible intervals for the $a$, $r$, and $b$ parameters. An additional option also allows the user to specify the model optimization period, where the time period can vary from a minimum of one day to the maximum of the entire length of the time series.  This differs from the BASEmetab approach where the optimization period is set at one day.  All results provided herein were generated using the EBASE R package.

@@

A preliminary assessment of the ability of EBASE to produce reasonable parameter estimates for $a$, $r$, and $b$ (@eq-ebase) was conducted using a simulated time series created with FWOXY. These assessments evaluated if 1) EBASE estimates similar values for the known parameters, and 2) how the estimated values change with different model configurations. The simulated time series used inputs from an approximate one year of continuous data from Apalachicola Bay, Florida.  The input data were 2012 observations from February 22nd to December 19th at the Cat Point water quality monitoring station (29.7021$^{\circ}$ N, -84.8802$^{\circ}$ W) and included 15 minute observations of dissolved oxygen (mg/L), water temperature (C), and salinity (psu).  Continuous meteorological observations from the nearby East Bay station (29.7909$^{\circ}$ N, -84.8834$^{\circ}$ W, ~10km north of Cat Point) were combined with the water quality data at the same time step and included air temperature (C), wind speed (m/s), and PAR (mmol/m$^2$ as total per 15 minute observation). An oxygen time series was generated from Fwoxy using the inputs from Apalachicola Bay, such that the simulated dissolved oxygen was created as close as possible to the observed given the inputs.  Known values for $a$ and $r$ parameters were returned for each time step of the simulated data, as well as all metabolic estimates for production, respiration, and gas exchange.  The only parameter that could not be simulated by Fwoxy was $b$, which was set to 0.251.

The simulated time series is shown in figure......

Several configurations of the EBASE model can influence how the metabolic rates are estimated. Two types of sensitivity analyses were conducted to evaluate changes in the parameters returned by EBASE, where in all cases, the parameters were compared to the known values in the simulated time series. First, parameter estimates were compared to the known values after changing characteristics of their prior distributions.  Second, parameter estimates were compared using a model period of optimization for one day and one week of observations.  Details of these analyses are as follows.

@@

The metabolic estimates returned from each analysis were compared to the known values to identify priors the model was most sensitive to and if the conclusions varied by optimization period.  Estimates for the modeled dissolved oxygen, gross production, respiration ($r$), gas exchange ($D$), and $a$ were compared to those from the simulated time series using coefficient of variation ($R^2$) and root mean square error ($RMSE$) summaries of the 1:1 comparison of the estimated vs simulated values.  For the one day optimization period, all parameter estimates returned at the time step of the input data (i.e., modeled dissolved oxygen, $D$, and $a$) were daily-averaged for comparison.  For the seven day optimization period, all parameter estimates and those from the known time series were weekly-averaged.  Estimates for the $b$ parameter were evaluated only by visual comparison because a single value was used in the simulated time series, preventing the estimate of quantitative comparisons.

Results for the sensitivity analyses are shown in ......

@@

Metabolic estimates from EBASE were further compared to those from existing methods to benchmark and evaluate potential differences from alternative model formulations used by each approach. The same one year continuous time series described above for water quality and weather data from Apalachicola Bay was used to compare metabolic estimates from EBASE, BASEmetab, and the Odum open-water method [@Odum56]. The Odum method used an approach described in @Murrell18 created for estuarine application that used a gas-exchange approach described in @Thebault08.  For better comparison with the other methods, the gas exchange parmaterization in the Odum method was modified to use the @Wanninkhof14 approach [as available in the WtRegDO R package, @Beck21]. The input data were provided to each method after converting the units as required (e.g., BASEmetab requires PAR as $\mu$mol m$^{-2}$ s$^{-1}$).  Water column depth at Cat Point is approximately 1.8 m and all daily metabolic estimates returned by each method were converted to volumetric rates by dividing by the depth ($\mu$mol O$_2$ m$^{-3}$ d$^{-1}$).  Each comparison also evaluated metabolic estimates using the observed dissolved oxygen time series and a detided time series using weighted regression [@Beck15].  The latter comparison provided an assessment of metabolic rates that are expected to be minimally influenced by tidal advection.  All metabolic estimates were obtained using the default settings for each of the three methods, with an exception that the prior distribution for the reaeration coefficient ($K$) in BASEmetab was informed by results from the Odum output.  Initial assessments indicated that unstable and inaccurate results are obtained from BASEmetab using an uninformed prior distribution for $K$.  All other prior distributions for BASEmetab and EBASE were set as the software default.  A one-day model optimization period was used for EBASE.

@@

### Application to representative NERR sites
